NFL Draft Analysis

### 2020-05-15 draft_analysis.Rmd
Updated prose

Load the NFL draft history data from *master_df*, containing the draft data from rounds 1-7 from the 1970-2016 NFL draft. The first line in this chunk will ensure that we will only keep the observations that do not have any missing data entries. Then, we will convert the necessary columns to integers and factors (rather than chars). This way, we can use the columns values to group and to produce regression lines in our data visualization steps later.

@@

Now let's process our 2020 Draft data. We will be using the *filter()* and *mutate()* methods found in the **dplyr** package in the **tidyverse** package. More info on **dplyr** can be found here: https://dplyr.tidyverse.org/

@@

We will use *mutate()* again to add a variable **period** for which period this entry is in.
Then for each period, we will take average career value of each position using *summarise()*, grouped by each round, period and position that the player was selected using *group_by()*. Both of these methods are also available in the **dplyr** package.

@@

@@

@@

Now let's focus our attention on the past 20 years and graph some visualizations of that data.

@@

@@

@@

qb_violin <- master_df %>%
  filter(year >= 2000) %>%
  filter(Pos == "QB") %>%
  ggplot(mapping=aes(x=factor(Rnd), y=CarAV)) +
    geom_violin()  +
    labs(x = "Rnd", y = "Career Value", title = "Average Value over Round for QBs Drafted in the 2000s") + 
    geom_smooth(method=lm)

qb_violin
```

@@

Next we will provide the same graphs to provide similar insight into the RB and WR selections of the draft. We will use *grid.arrange* to show the graphs side by side in three columns.

@@

Not as drastic of a difference in rounds 1 and 2 in the WR group. The RB average value in the 2000s seems to resemble the QB position.

@@

Much less steep of a drop off between round 1 to 2 in the WR group.

Again we see a trend that by pick 100 the value seems to hit a plateue.

@@

qb_violin <- master_df %>%
  filter(year >= 2000) %>%
  filter(Pos == "QB") %>%
  ggplot(mapping=aes(x=factor(Rnd), y=CarAV)) +
    geom_violin()  +
    labs(x = "Rnd", y = "Career Value", title = "QB Average Value") + 
    geom_smooth(method=lm)

rb_violin <- master_df %>%
  filter(year >= 2000) %>%
  filter(Pos == "RB") %>%
  ggplot(mapping=aes(x=factor(Rnd), y=CarAV)) +
    geom_violin()  +
    labs(x = "Rnd", y = "Career Value", title = "RB Average Value") + 
    geom_smooth(method=lm)

wr_violin <- master_df %>%
  filter(year >= 2000) %>%
  filter(Pos == "WR") %>%
  ggplot(mapping=aes(x=factor(Rnd), y=CarAV)) +
    geom_violin()  +
    labs(x = "Rnd", y = "Career Value", title = "WR Average Value") + 
    geom_smooth(method=lm)

@@

RBs in Round 1 always produce.
We again see that round 2 is a value round for WRs as it appears to resemble the type of value you get out of round 1.

@@

First, lets control the resampling of our data. In this case, we’re going to cross-validate the data 3 times, therefore training the model 3 times on different portions of the data before settling on the best tuning parameters. For gbm, the tuning parameters are trees, shrinkage, and interaction depth.

@@

objControl <- trainControl(method='cv', number=3, returnResamp='none', summaryFunction = twoClassSummary, classProbs = TRUE)

@@

First, we will set the predictor variables to **Tm**, **Pos**, and **Pick**, and the sole outcome variable as **roty**. Then we will use the *train()* method from the **caret** package to train the model using the *off_df* variable that contains all our historical draft data on all the offensive skill position players. We tell *train()* what predictor and outcome variables to use, as well as to use the gbm method with the resampling control that we set previously.

@@

Now let's get the top 10 players that the model is most certain will be the Offensive Rookie of the Year for 2017, 2018, and 2019 respectively.

@@

**Joe Burrow is our model's top prediction to win the MVP!!!** We also see that Cam Akers, Tua, Henry Ruggs III, Jerry Jeudy, CeDee Lamb, and Justin Jefferson also appear in the top 10. All of these players are among those that have the highest odds to win the the Offensive MVP according to OddsShark.

>>>>

First, we will tidy the *master_df* dataframe. The first line in this chunk will ensure that we will only keep the observations that do not have any missing data entries. Then, we will convert the necessary columns to integers and factors (rather than chars). This way, we can use the columns values to group and to produce regression lines in our data visualization steps later.

@@

Now let's process/tidy our 2020 Draft data. We will be using the *filter()* and *mutate()* methods found in the **dplyr** package in the **tidyverse** package. More info on **dplyr** can be found here: https://dplyr.tidyverse.org/

@@

We will use *mutate()* again to add a variable **period** for which period this entry is in.
Then for each grouping of period, round, and period using *group_by()*, we will create an average career value variable  **Avg_AV** of each grouping using *summarise()*. Both of these methods are also available in the **dplyr** package.

@@

Here, we map the x-variable to **year** and the y-variable to **Avg_AV**, and using *geom_point()* tells **ggplot2** to create a scatter plot of the data. We them use *geom_smooth(method=lm)* to create a regression line, where lm stands for linear model. More info can be found here: https://ggplot2.tidyverse.org/reference/geom_smooth.html

We group by round to create 7 different regression lines in our plot.

@@

To create a bar graph, we use *geom_bar*, where stat="identity" allows us to have the height of the bars represent values in the data, which are mapped to **Avg_AV** in this case. The position="dodge" puts the bars for each round side by side.

More info on *geom_bar()*: https://ggplot2.tidyverse.org/reference/geom_bar.html

@@

Now let's focus our attention on QBs of the past 20 years and graph some visualizations of that data. To do this, we will use *filter()* on **Pos** and **year**.

@@

In this case, in addition to the linear model regression line in green, we will also add a locally estimated scatterplot smoothing (LOESS) regression line using *geom_smooth(method='loess')*.

@@

We use *geom_violion()* to create the violin plots, more info can be found here: https://ggplot2.tidyverse.org/reference/geom_violin.html

@@

qb_violin <- master_df %>%
  filter(year >= 2000) %>%
  filter(Pos == "QB") %>%
  ggplot(mapping=aes(x=factor(Rnd), y=CarAV)) +
    geom_violin()  +
    labs(x = "Rnd", y = "Career Value", title = "Average Value over Round for QBs Drafted in the 2000s")
qb_violin
```

@@

Next we will provide the same graphs to provide similar insight into the RB and WR selections of the draft. We will use *grid.arrange* to show the graphs for WRs, RBs, and QBs side by side in three columns.

@@

Insights:
Not as drastic of a difference in rounds 1 and 2 in the WR group. 
The RB average value in the 2000s seems to resemble the QB position.

@@

Insights:
Much less steep of a drop off between round 1 to 2 in the WR group.
Again we see a trend that by pick 100 the value seems to hit a plateue.

@@

qb_violin <- master_df %>%
  filter(year >= 2000) %>%
  filter(Pos == "QB") %>%
  ggplot(mapping=aes(x=factor(Rnd), y=CarAV)) +
    geom_violin()  +
    labs(x = "Rnd", y = "Career Value", title = "QB Average Value")

rb_violin <- master_df %>%
  filter(year >= 2000) %>%
  filter(Pos == "RB") %>%
  ggplot(mapping=aes(x=factor(Rnd), y=CarAV)) +
    geom_violin()  +
    labs(x = "Rnd", y = "Career Value", title = "RB Average Value")

wr_violin <- master_df %>%
  filter(year >= 2000) %>%
  filter(Pos == "WR") %>%
  ggplot(mapping=aes(x=factor(Rnd), y=CarAV)) +
    geom_violin()  +
    labs(x = "Rnd", y = "Career Value", title = "WR Average Value")

@@

Insights:
RBs in Round 1 always produce, and RBs selected in later rounds have a much larger spread than quarterbacks, suggesting it is easier to get a running back of decent value in later rounds than it is to get a QB.
We again see that round 2 is a value round for WRs as it appears to resemble the type of value you get out of round 1. Rounds 1-4 have generally similar range in the spread of values, so there seems to be great talent available outside of rounds 1 and 2 for wide receivers.

@@

First, lets control the resampling of our data. In this case, we’re going to cross-validate the data 3 times, so we will train the model 3 times on different portions of the data before deciding on the best tuning parameters. For the gbm method that we will be using, the tuning parameters are trees, shrinkage, and interaction depth.

@@

objControl <- trainControl(method='cv', number=3, returnResamp='none', summaryFunction = twoClassSummary, classProbs = TRUE)
```

@@

First, we will set the predictor variables to **Tm**, **Pos**, and **Pick**, and the sole outcome variable as **roty**. Then we will use the *train()* method from the **caret** package to train the model using the *off_df* variable that contains all our historical draft data on all the offensive skill position players. We tell *train()* what predictor and outcome variables to use, as well as to use the gbm (gradient boosting algorithm) with the resampling control that we set previously.

@@

So we see that the model is 85% percent sure that Mitchell Trubisky will not win Rookie Offensive MVP, and ~15% sure that Trubisky will win the award.

Now let's get the top 10 players that the model is most certain will be the Offensive Rookie of the Year for 2017, 2018, and 2019 respectively, using the *top_n()* function in the **dplyr** package (https://dplyr.tidyverse.org/reference/top_n.html)

@@

**Joe Burrow is our model's top prediction to win the MVP!!!** The model is 21% sure that Burrow will win the Rookie Offensive MVP, and 79% sure that he will not.

We also see that Cam Akers, Tua, Henry Ruggs III, Jerry Jeudy, CeDee Lamb, and Justin Jefferson also appear in the top 10. All of these players are among those that have the highest odds to win the the Offensive MVP according to OddsShark.
